---
title: "Calibrating the detectors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating the detectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(staleness)
library(metafor)
```

Implementing five published detectors faithfully, as
`vignette("methods", package = "staleness")` does, answers the easy half of
the problem. The hard half is: given a meta-analysis today, which of the five
should you actually trust? This vignette is the scientific argument for why
that question cannot be answered by reading the methods papers, and for how
this package answers it instead — with data.

## 1. The problem

The five methods have been run side by side exactly once, in the comparative
evaluation this package cites throughout (Pattanittum et al., 2012, applying
them to 80 published systematic reviews). Their agreement with each other was
**Kappa = 0.14** — indistinguishable from chance, and computed over the three
methods that discriminated at all: `ottawa` flagged 34 of the 80 reviews,
`rcma` flagged 7 and `barrowman` flagged 7. The other two, `sufficiency` and
`simulation`, flagged nothing whatsoever. Only one review was flagged by all
three of the discriminating methods at once.

That is not a story about one weak method among four strong ones. It is a
story about nobody knowing, in the thirteen years since, which of the five
methods to reach for — because no software existed that could run all five
against real history and check which ones were actually right.

## 2. The circularity trap

The obvious next step is to pick a definition of "was this meta-analysis
really out of date at time t?" and score every detector's historical
verdicts against it. The obvious definition — "the conclusion changed:
significance flipped, or the effect reversed sign" — is a trap.

`ottawa()` signals, in part, on exactly that: a change in statistical
significance. If "truth" is defined the same way, `ottawa` is correct by
construction, for every review, regardless of whether the method has any
real predictive value. Scoring it against that truth would not be measuring
calibration. It would be measuring whether `ottawa()`'s source code agrees
with itself, dressed up as an empirical result.

This package resolves that by never collapsing "truth" into a single
definition, and by tracking which detector-truth pairs share enough logic to
be circular, in the data itself rather than in a footnote someone can miss:

```{r contaminated-pairs}
CONTAMINATED_PAIRS
```

## 3. Three independent truths

Every backtest is scored under three separate definitions of "the
meta-analysis changed", each implemented in `R/truth.R` and listed by
`available_truths()`:

| Truth | What it measures |
|---|---|
| `shift` | How far the pooled estimate at time `t` sits from the final estimate, judged against how precisely we know the final estimate today (`|theta_final - theta_t| / SE_final`). |
| `surprise` | The same distance, but judged against how precisely the estimate was known *at the time* (`|theta_final - theta_t| / SE_t`) — how surprising the eventual result would have looked to someone standing at `t`. |
| `conclusion` | Whether the practical conclusion changed: a flip in the sign of the effect, or a flip in statistical significance. |

`shift` and `surprise` never look at significance thresholds or effect-size
ratios, so no detector's own logic can leak into them. `conclusion` is the
one exception, and only for `ottawa`, which is exactly why it is the pair
`CONTAMINATED_PAIRS` names.

## 4. The contamination declaration

`calibration()` computes sensitivity, specificity and false-alarm rate for
every detector against a chosen truth, and it never lets a reader miss which
rows are circular: every row carries a `contaminated` column, looked up from
`CONTAMINATED_PAIRS`, sitting right next to the numbers it warns about.

```{r contaminated-mark}
dat <- metadat::dat.bcg
es  <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg,
              data = dat)
ma  <- rma(yi, vi, data = es)
stream <- evidence_stream(ma, date = es$year, study_id = seq_along(es$year))

bt <- backtest(stream, cuts = "yearly", horizon = 3, window = 5,
              min_k = 3, seed = 42)
calibration(bt, "conclusion")
```

`ottawa`'s row above reads `contaminated = TRUE`. Nothing stops a caller from
using that number — sometimes the circular comparison is exactly what you
want, if what you are asking is "does `ottawa` reproduce its own rule
correctly" rather than "does `ottawa` predict the future" — but nobody
reading this table can mistake it for an independent result. `rcma` and
`sufficiency`, scored against the same truth, are not marked, because
neither shares `conclusion`'s specific logic.

A relationship between two *detectors* deserves the same care, and earlier
versions of this vignette got one wrong. `rcma()` and `ottawa()` look like the
same effect rule at the same thresholds, and on mean differences they are. On
ratio measures they are not: `ottawa()` compares relative risk *reductions*,
`(1 - RR_new) / (1 - RR_prev)`, which is what its published application
specifies. Of the ten reviews with the largest Ottawa indicator in that study,
all ten fire under that definition and none fires under the ratio of the
effects. So neither detector's firings are a subset of the other's, and the
containment this vignette used to declare was an artefact of implementing both
criteria with one function. See `?ottawa` and
`vignette("methods", package = "staleness")` for the measured comparison.

## 5. A backtest, executed

The backtest above walks `stream` one year at a time. At each cut it
refits the meta-analysis using only studies published on or before that
year (never anything later — this is checked by a dedicated test that
verifies no detector can fire before a synthetic break actually happens),
asks every detector the question it would be asked today, and compares the
verdict with what the complete series eventually showed.

```{r backtest-summary}
bt
summary(bt)
```

Four things are worth noticing in this particular run before reading
anything into the sensitivity and specificity numbers.

First, `barrowman` and `simulation` carry `n = 0` and `NA` metrics in every
row. Both require a prior meta-analysis that was *not yet statistically
significant*, and in this series — a strong, consistent protective effect of
BCG vaccination — even the earliest snapshots were already significant at
conventional thresholds. Both methods are `not_applicable` at every single
cut, so there is nothing to score. `calibration()` reports that as a row
rather than as an absence: "this detector never applied to this evidence" is
itself a result about the detector, and a table that simply omitted the two
would leave a reader to notice the gap, or not. It is the narrow-scope
critique from `vignette("methods", package = "staleness")`, visible directly
in real data rather than argued in the abstract. Note also the `n` column:
`sufficiency` scores 18 cuts where `rcma` and `ottawa` score 19, because at
the earliest cut the updated meta-analysis still had fewer than five studies
and `sufficiency()` declined that one too.

Second, this run uses `horizon = 3` with `window = 5`, and the censoring
follows the *larger* of the two. A cut needs three years of future for its
truth to have materialised, but it also needs five years of future for the
detector to be handed the same amount of new evidence every other scored cut
was handed. Censoring on the horizon alone would leave the last cuts looking
evaluable while their five-year evidence window ran off the end of the data —
they would see one or two new studies instead of a full window, be shown less
change than really happened, and be scored as full observations anyway.
That biases exactly those cuts toward `current`, invisibly. `backtest()`
censors on `max(horizon, window)` for that reason, which is why the
`censored` count above is larger than the horizon alone would produce.

Third, `sufficiency` is the only detector here with a specificity below 1:
it fires once, at the 1955–1979 cut for 1971, on a year with no true event
under any of the three truth definitions. That single firing is not a bug,
and it is worth being precise about why. `rcma()` and `ottawa()` are
deterministic threshold rules — they fire when a fixed quantity crosses a
fixed line, so on evidence that is not moving they never fire at all, and a
specificity of exactly 1 costs them nothing. `sufficiency()`'s stability half
is a permutation test at `alpha_stability = 0.05`, and an honest test at the
5% level is *entitled* to one false positive in twenty; over 18 scored cuts,
one is the modal outcome. A version of it with specificity 1 on this series
would be evidence that the test had no power, not that it was better — which
is exactly what the previous implementation turned out to be, silent on all
23 cuts and, as `vignette("methods")` records, blind to any change confined
to the last few studies of a series. Reading these two columns as
commensurable across the five methods is a mistake the underlying literature
makes routinely; a threshold rule and a level-alpha test are not the same
kind of object, and only one of them has a false-alarm rate you can set.

Fourth, this is thirteen studies over three decades: enough to demonstrate
the machinery end to end, nowhere near enough to draw a general conclusion
about which detector deserves trust across the literature. `backtest()`
works the same way regardless of series length; drawing a real verdict about
the five methods needs many series like this one, not one.

```{r plot-calibration, fig.width = 6, fig.height = 4, fig.alt = "Bar chart of sensitivity and specificity per detector, scored against the shift truth definition"}
plot(bt, truth = "shift")
```

## 6. Lead time

Sensitivity and specificity, on their own, cannot tell a genuinely useful
early-warning detector from one that only ever notices a change after the
evidence has already moved. A detector that fires the same year the shift
happens looks identical, in a contingency table, to one that fired five
years ahead of it — both are "hits". In practice only one of them is worth
having: an alarm that never rings before the fire has already spread saves
nobody.

`lead_time()` reports, per detector, the median number of years between a
detector's earliest correct `out_of_date` firing and the true event it
caught (per whichever truth definition is passed in), across every event the
detector caught at all:

```{r lead-time}
lead_time(bt, truth = "shift")
```

Every `median_lead` above is `NA`, not zero, and the two reasons are
different. For `rcma`, `ottawa` and `sufficiency`, `n_events = 5`: five true
shifts happened in this series, per the `shift` definition, and none of the
three applicable detectors ever flagged one of them as `out_of_date` —
neither early nor late, which is the sensitivity 0 visible in `summary(bt)`.
`sufficiency`'s one firing does not change that: it landed on a cut with no
true event, which is why it shows up in the specificity column and not here.
`lead_time()` reports `NA` rather than inventing a number for events that
were never caught at all; a real numeric lead only exists for events a
detector did catch. For `barrowman` and `simulation`, `n_events = 0`: they
were `not_applicable` at every cut, so there were no eligible rows in which
an event could even be located. The distinction between "had five chances and
took none" and "never got a chance" is exactly what the two columns together
make visible, and why a method with nothing to score still gets a row.

That distinction matters because a numeric `median_lead` of exactly zero
would describe a *different*, and in one sense better, failure: a detector
that does catch every event, but only ever in the same period the evidence
had already moved. Sensitivity and specificity cannot tell that detector
apart from one with genuine foresight — both score as perfect hits in a
contingency table. Only `lead_time()` can, and a high-sensitivity detector
with a `median_lead` of zero is not a useful early-warning system: it is a
slower way of noticing something a reader would have seen anyway by looking
at the updated forest plot. Distinguishing "never caught it" from "caught it
too late to matter" from "caught it early enough to act on" is the one
distinction this package's backtesting adds that the underlying literature
has never reported, and it is what actually decides whether calibrating
these detectors is worth doing at all.
