---
title: "Calibrating the detectors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating the detectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(staleness)
library(metafor)
```

Implementing five published detectors faithfully, as
`vignette("methods", package = "staleness")` does, answers the easy half of
the problem. The hard half is: given a meta-analysis today, which of the five
should you actually trust? This vignette is the scientific argument for why
that question cannot be answered by reading the methods papers, and for how
this package answers it instead — with data.

## 1. The problem

The five methods have been run side by side exactly once, in the comparative
evaluation this package cites throughout (Pattanittum et al., 2012, applying
them to 80 published systematic reviews). Their agreement with each other was
**Kappa = 0.14** — indistinguishable from chance. Of the three methods that
produced usable verdicts on that sample, `ottawa` flagged 34 reviews as out
of date, `barrowman` flagged 7, and the simulation method flagged none. Only
one review was flagged by all three at once.

That is not a story about one weak method among four strong ones. It is a
story about nobody knowing, in the thirteen years since, which of the five
methods to reach for — because no software existed that could run all five
against real history and check which ones were actually right.

## 2. The circularity trap

The obvious next step is to pick a definition of "was this meta-analysis
really out of date at time t?" and score every detector's historical
verdicts against it. The obvious definition — "the conclusion changed:
significance flipped, or the effect reversed sign" — is a trap.

`ottawa()` signals, in part, on exactly that: a change in statistical
significance. If "truth" is defined the same way, `ottawa` is correct by
construction, for every review, regardless of whether the method has any
real predictive value. Scoring it against that truth would not be measuring
calibration. It would be measuring whether `ottawa()`'s source code agrees
with itself, dressed up as an empirical result.

This package resolves that by never collapsing "truth" into a single
definition, and by tracking which detector-truth pairs share enough logic to
be circular, in the data itself rather than in a footnote someone can miss:

```{r contaminated-pairs}
CONTAMINATED_PAIRS
```

## 3. Three independent truths

Every backtest is scored under three separate definitions of "the
meta-analysis changed", each implemented in `R/truth.R` and listed by
`available_truths()`:

| Truth | What it measures |
|---|---|
| `shift` | How far the pooled estimate at time `t` sits from the final estimate, judged against how precisely we know the final estimate today (`|theta_final - theta_t| / SE_final`). |
| `surprise` | The same distance, but judged against how precisely the estimate was known *at the time* (`|theta_final - theta_t| / SE_t`) — how surprising the eventual result would have looked to someone standing at `t`. |
| `conclusion` | Whether the practical conclusion changed: a flip in the sign of the effect, or a flip in statistical significance. |

`shift` and `surprise` never look at significance thresholds or effect-size
ratios, so no detector's own logic can leak into them. `conclusion` is the
one exception, and only for `ottawa`, which is exactly why it is the pair
`CONTAMINATED_PAIRS` names.

## 4. The contamination declaration

`calibration()` computes sensitivity, specificity and false-alarm rate for
every detector against a chosen truth, and it never lets a reader miss which
rows are circular: every row carries a `contaminated` column, looked up from
`CONTAMINATED_PAIRS`, sitting right next to the numbers it warns about.

```{r contaminated-mark}
dat <- metadat::dat.bcg
es  <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg,
              data = dat)
ma  <- rma(yi, vi, data = es)
stream <- evidence_stream(ma, date = es$year)

bt <- backtest(stream, cuts = "yearly", horizon = 3, window = 5,
              min_k = 3, seed = 42)
calibration(bt, "conclusion")
```

`ottawa`'s row above reads `contaminated = TRUE`. Nothing stops a caller from
using that number — sometimes the circular comparison is exactly what you
want, if what you are asking is "does `ottawa` reproduce its own rule
correctly" rather than "does `ottawa` predict the future" — but nobody
reading this table can mistake it for an independent result. `rcma` and
`sufficiency`, scored against the same truth, are not marked, because
neither shares `conclusion`'s specific logic.

## 5. A backtest, executed

The backtest above walks `stream` one year at a time. At each cut it
refits the meta-analysis using only studies published on or before that
year (never anything later — this is checked by a dedicated test that
verifies no detector can fire before a synthetic break actually happens),
asks every detector the question it would be asked today, and compares the
verdict with what the complete series eventually showed.

```{r backtest-summary}
bt
summary(bt)
```

Two things are worth noticing in this particular run before reading anything
into the sensitivity and specificity numbers. First, `barrowman` and
`simulation` do not appear at all: both require a prior meta-analysis that
was *not yet statistically significant*, and in this series — a strong,
consistent protective effect of BCG vaccination — even the earliest
two-study snapshot, from 1949, was already significant at conventional
thresholds. Both methods are `not_applicable` at every single cut, which
`calibration()` correctly excludes rather than silently scoring as
agreement. That is the narrow-scope critique from
`vignette("methods", package = "staleness")`, visible directly in real data
rather than argued in the abstract.

Second, this is thirteen studies over three decades: enough to demonstrate
the machinery end to end, nowhere near enough to draw a general conclusion
about which detector deserves trust across the literature. `backtest()`
works the same way regardless of series length; drawing a real verdict about
the five methods needs many series like this one, not one.

```{r plot-calibration, fig.width = 6, fig.height = 4, fig.alt = "Bar chart of sensitivity and specificity per detector, scored against the shift truth definition"}
plot(bt, truth = "shift")
```

## 6. Lead time

Sensitivity and specificity, on their own, cannot tell a genuinely useful
early-warning detector from one that only ever notices a change after the
evidence has already moved. A detector that fires the same year the shift
happens looks identical, in a contingency table, to one that fired five
years ahead of it — both are "hits". In practice only one of them is worth
having: an alarm that never rings before the fire has already spread saves
nobody.

`lead_time()` reports, per detector, the median number of years between a
detector's earliest correct `out_of_date` firing and the true event it
caught (per whichever truth definition is passed in), across every event the
detector caught at all:

```{r lead-time}
lead_time(bt, truth = "shift")
```

Every `median_lead` above is `NA`, not zero, and the reason is visible in
`summary(bt)`: `rcma`, `ottawa` and `sufficiency` all had sensitivity 0
against `shift` in this run, with `n_events = 5` in every row. Five true
shifts happened in this series, per the `shift` definition, and none of the
three applicable detectors ever flagged one of them as `out_of_date` —
neither early nor late. `lead_time()` reports `NA` rather than inventing a
number for events that were never caught at all; a real numeric lead only
exists for events a detector did catch.

That distinction matters because a numeric `median_lead` of exactly zero
would describe a *different*, and in one sense better, failure: a detector
that does catch every event, but only ever in the same period the evidence
had already moved. Sensitivity and specificity cannot tell that detector
apart from one with genuine foresight — both score as perfect hits in a
contingency table. Only `lead_time()` can, and a high-sensitivity detector
with a `median_lead` of zero is not a useful early-warning system: it is a
slower way of noticing something a reader would have seen anyway by looking
at the updated forest plot. Distinguishing "never caught it" from "caught it
too late to matter" from "caught it early enough to act on" is the one
distinction this package's backtesting adds that the underlying literature
has never reported, and it is what actually decides whether calibrating
these detectors is worth doing at all.
