---
title: "The five detectors and what is wrong with them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The five detectors and what is wrong with them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(staleness)
library(metafor)
```

Five statistical methods for deciding whether a meta-analysis needs updating
were published between 2003 and 2014. None of them ever received a reusable
software implementation, so nobody has ever been able to run all five against
the same body of evidence and see where they agree. This package implements
all five as pure functions with a common signature — `(prev, new_ma, ...)` in,
a `staleness_verdict` out — precisely so that can finally be done (see
`vignette("backtesting", package = "staleness")` for what happens when it is).

This vignette is deliberately unflattering. Each section below states a
detector's formula, where it comes from, and the specific way the published
literature — or this package's own backtesting — shows it can mislead you.
Implementing a method faithfully is not the same as endorsing it.

## `rcma()`: recursive cumulative meta-analysis

```
ratio = theta_updated / theta_prior      signal if ratio <= 0.5 or ratio >= 1.5
```

The idea, due to Ioannidis and Lau's work on how treatment effects evolve as
evidence accumulates (*Proceedings of the National Academy of Sciences*,
2001), is the simplest of the five: refit the meta-analysis after each new
study and watch how far the pooled estimate moves. A move of 50% or more in
either direction is the signal.

```{r rcma-signal}
prev   <- rma(yi = rep(log(0.50), 4), vi = rep(0.04, 4), measure = "RR")
new_ma <- rma(yi = rep(log(0.20), 4), vi = rep(0.04, 4), measure = "RR")
rcma(prev, new_ma)
```

**What is wrong with it.** The ratio is only well-behaved for measures that
are themselves ratios (risk ratio, odds ratio, hazard ratio): those live on a
log scale where `exp(a - b)` is a well-defined, symmetric ratio regardless of
how small the effect is. For difference measures — a mean difference, a raw
risk difference — the ratio is undefined once the prior effect is close to
zero: a prior estimate of 0.02 moving to 0.40 produces a ratio of 20, an
artefact of the denominator, not a real twenty-fold change. The method as
published does not address this case at all.

```{r rcma-na}
prev_md <- rma(yi = rep(0.02, 4), vi = rep(0.25, 4), measure = "MD")
new_md  <- rma(yi = rep(0.40, 4), vi = rep(0.25, 4), measure = "MD")
rcma(prev_md, new_md)
```

This package's answer is to read the measure straight off the `rma` object
and, for difference measures with a prior effect indistinguishable from zero,
return `not_applicable` with the reason attached rather than a number that
looks plausible and is not.

## `ottawa()`: the Ottawa method

Two quantitative signals: a change in statistical significance (using
**p < 0.04**, not the conventional 0.05 — this is the threshold in the
original method, not a typo), and a change in the pooled effect of at least
50%, computed the same way `rcma()` computes it. The method is due to Shojania
and colleagues' survival analysis of how quickly systematic reviews go out of
date (*Annals of Internal Medicine*, 2007), the same paper that established
the headline numbers this package's introduction vignette opens with.

```{r ottawa-signal}
prev_o <- rma(yi = c(0.02, -0.01, 0.03, -0.02), vi = rep(0.05, 4), measure = "RR")
new_o  <- rma(yi = rep(log(0.40), 8), vi = rep(0.01, 8), measure = "RR")
ottawa(prev_o, new_o)
```

The original method also lists four *qualitative* signals: opposing
findings, evidence of substantial harm, a new superior therapy, and a change
in the balance of effectiveness. **This package does not infer any of
them.** They are accepted as a character vector supplied by the analyst
(`qualitative = "substantial harm reported"`, for instance) and reported
separately from the two quantitative signals. An algorithm that claimed to
judge "substantial harm" from effect sizes alone would be a much bigger
promise than the data can support, and would spend the credibility of the
whole package to make it.

**Declared: `ottawa` contains `rcma` verbatim.** The effect-change half of
the Ottawa method *is* `rcma()`'s entire rule — the same ratio, computed by
the same internal function, against the same 0.5 and 1.5 thresholds. At
default parameters, therefore:

```
rcma(prev, new_ma) == "out_of_date"   implies   ottawa(prev, new_ma) == "out_of_date"
```

always, by construction. Verified over 400 random prior/updated pairs:
`rcma` fired 63 times, `ottawa` fired 116 times, and every one of `rcma`'s 63
was among `ottawa`'s — zero disjoint firings, as the algebra requires.

This is faithful to the published methods and is not a defect in either: the
Ottawa method genuinely adopts the rCMA effect-change criterion and adds a
significance criterion and four qualitative ones on top of it. It is declared
here for the same reason `CONTAMINATED_PAIRS` exists. This package goes to
some trouble to make sure nobody mistakes a structural identity between a
detector and a truth definition for an empirical finding; a structural
dependency between two *detectors* deserves the same treatment. The headline
question this package exists to re-ask — the five methods agreed at
Kappa = 0.14 on 80 reviews — involves ten detector pairs, and one of them is
deterministic. Anyone recomputing that agreement should be counting nine.

## `barrowman()`: the participant-ratio method

```
q = m / n            n = (Z_crit^2 * N) / Z^2            signal if q > 1
```

`m` is the number of participants contributed by the new studies; `N` is the
total in the prior meta-analysis; `Z` is that meta-analysis's own test
statistic; `Z_crit` is 1.96. `n` is a back-calculation: how many new
participants, added at the same average effect and variance as the existing
evidence, would it take to push the pooled result across the significance
threshold. `q > 1` means the new evidence already supplied more than that.

```{r barrowman-signal}
prev_b <- rma(yi = c(0.10, -0.05, 0.08, -0.02), vi = rep(0.05, 4), measure = "MD")
z <- prev_b$zval
n_req <- (1.96^2 * 400) / z^2
barrowman(prev_b, n_prev = 400, n_new = ceiling(n_req * 1.5))
barrowman(prev_b, n_prev = 400, n_new = floor(n_req * 0.5))
```

**What is wrong with it.** The method, proposed by Barrowman and colleagues
as a way to identify null meta-analyses ripe for updating (*BMC Medical
Research Methodology*, 2003), only makes sense when the prior meta-analysis
was *not* statistically significant — `n` is derived from how far the prior
result sits from the significance boundary, and that derivation has no
answer when the prior result was already on the far side of it. This
restricts the method to a narrow slice of published meta-analyses. When it
does not apply, `barrowman()` returns `not_applicable`, never `current`:
treating "the method has nothing to say" as "the review is fine" would hand
this detector an unearned advantage in any calibration exercise, since it
would then never be wrong on exactly the cases it declined to judge.

```{r barrowman-na}
sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
barrowman(sig, n_prev = 400, n_new = 900)
```

## `sufficiency()`: sufficiency and stability

```
index = N_fs / (5k + 10)      stability = slope of the cumulative pooled effect
```

`N_fs` is Rosenthal's fail-safe N: an estimate of how many unpublished null
studies it would take to erase a significant pooled effect. A review is
*sufficient* when the index exceeds 1 — Rosenthal's own rule of thumb for a
result being robust to unpublished null findings. It is *stable* when the
slope of the cumulative pooled effect, plotted study by study in publication
order, is no larger than the same studies would produce in a random order:
the estimate has stopped moving as evidence accrued. That last clause is a
declared departure from the published rule, and the next-but-one subsection
sets out exactly what was changed and why.

The method is due to Mullen, Muellerleile and Bryant's treatment of
cumulative meta-analysis indicators (*Personality and Social Psychology
Bulletin*, 2001). This package follows the specific two-snapshot form of it
used in the comparative evaluation of Pattanittum and colleagues (*PLoS
ONE*, 2012), because that is the only source that defines sufficiency and
stability as a comparison between a *prior* meta-analysis and an *updated*
one, rather than as a property of a single, static body of studies.

**The rule, and why it is easy to get backwards.** A review counts as
out of date only when it is **both sufficient and unstable at once**:

```{r sufficiency-out}
yi <- seq(log(0.9), log(0.2), length.out = 20)
vi <- rep(0.01, 20)
drifting <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(drifting, drifting)
```

Insufficiency by itself — an index at or below 1 — is **never** grounds for
"out of date", no matter how unstable the cumulative effect looks:

```{r sufficiency-insufficient}
yi <- rep(log(0.99), 6)
vi <- rep(0.04, 6)
weak <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(weak, weak)
```

This is the primary source's own reading, confirmed against its data, not
just its prose: of the 80 reviews in the Pattanittum et al. comparison, all
80 had a fail-safe ratio below 1 (insufficient), and for exactly that reason
the method could not flag a single one of them as out of date. Had the
opposite direction been correct — insufficiency implying "out of date" — that
same fact about the 80 reviews would have produced the opposite result: all
80 flagged.

The two halves also read from different bodies of evidence, and that is
deliberate, not an oversight: sufficiency — the fail-safe N and the `k` in
`5k + 10` — is computed from the meta-analysis **as it stood when
previously published**; stability is computed from the cumulative series of
the **updated** meta-analysis, because that is where drift from new studies
would actually show up. See section 4.4 of the package's design document for
the full paper trail behind this, including the case that would silently
give the wrong verdict if both halves were read from the same snapshot.

### Declared adaptation: how stability is actually tested

Every other adaptation in this package is documented at length — `rcma()`'s
`not_applicable` rule for difference measures, `ottawa()`'s refusal to infer
qualitative signals, the prev/updated split above, the inverted direction of
the out-of-date criterion. This one is the last of them, and it matters most,
because the backtesting results will be read as findings about the published
method.

**What the source says.** Pattanittum et al. (2012, Table 1), reproducing
Mullen et al.'s rule, states the out-of-date criterion as *"failsafe ratio >1
and absolute slope of the linear regression >0"*.

**Why it cannot be implemented as written.** On continuous data the
least-squares slope of a cumulative series is never exactly zero. Taken
literally, `|slope| > 0` is true of essentially every real meta-analysis, so
the stability indicator would carry no information at all and the method
would reduce to "sufficient ⇒ out of date". Some rule for deciding when a
slope is *large enough* has to stand in for it.

**What was tried first, and why it was wrong.** The obvious substitute is the
t-test on the slope that `lm()` reports, treating `|slope| > 0` as "slope
significantly different from zero". That test is invalid here, and
measurably so. A cumulative weighted mean is near-perfectly autocorrelated by
construction, and it converges on the pooled effect by the law of large
numbers whether or not anything is changing, so the OLS null distribution
does not apply: the test detects *convergence* and reports *instability*.
Measured over 300 samples of evidence that genuinely does not change — 20
prior studies and 10 new ones drawn from one distribution — it returned
`out_of_date` **209 times**. `rcma()` and `ottawa()` returned it zero times
on the same 300 samples. At the degenerate limit it was worse: twelve
byte-identical studies give a cumulative series spanning 2.2e-16, and the fit
obligingly returned a slope of -7.4e-17 with p = 0.014, i.e. "out of date".

**What runs instead.** A **permutation test over study order**. The observed
slope is compared with the slopes obtained by reshuffling the same studies
999 times and recomputing the cumulative series each time; `p_slope` is the
two-sided proportion of permuted slopes at least as large in absolute value,
and `alpha_slope` (default 0.05) is the cutoff. A cumulative series that is
constant to within floating-point rounding is short-circuited as maximally
stable before anything is fitted.

Three reasons this is the right substitute. It assumes nothing about the
independence of a series that is autocorrelated by design. It asks the
scientifically meaningful question directly — *is the drift in the cumulative
effect larger than what these same studies would produce in a random
order?* — which is a sharper reading of "the estimate has stopped moving"
than the source's own wording. And because study order is exchangeable when
evidence is not changing, the test is exact under that null, so the
package's stated invariant (statistically identical new evidence raises no
signal) holds by construction rather than by luck. On the same 300-sample
experiment the permutation version returns `out_of_date` **12 times** — the
nominal 5%, within sampling error, instead of 70%.

The `slope` itself is still reported in `detail`, unchanged and on its
original scale, so anyone who wants to apply the literal published rule can
read it straight off the verdict.

**What is wrong with it, independent of any of the above.** Rosenthal's
fail-safe N has been discredited as a measure of publication bias since
Becker (2005) — it answers a narrower and less useful question than it is
usually taken to answer, and better alternatives (trim-and-fill, funnel-plot
regression) have existed for years. `sufficiency()` implements it
anyway, faithfully, exactly as the published method specifies it, because
this package's purpose is not to pre-judge which of the five methods
deserves to survive contact with data. If Rosenthal's N is as weak a
foundation as the critique suggests, the backtesting engine will show it —
that is a fact to be established empirically, in
`vignette("backtesting", package = "staleness")`, and not decided in advance
by leaving the method out.

## `simulation()`: prospective power

Ten thousand replicate "new studies" are simulated using the parameters of
the recently published literature, added to the prior meta-analysis, and the
fraction of replicates in which the pooled result reaches significance is
recorded as a power estimate. A power at or above 80% is the signal.

```{r simulation-signal}
prev_sim <- rma(yi = c(0.30, 0.28, 0.32, 0.29), vi = rep(0.16, 4), measure = "MD")
new_ev   <- list(yi = c(0.30, 0.31, 0.29), vi = rep(0.005, 3), k = 3)
simulation(prev_sim, new_ev, seed = 3)
```

Like `barrowman()`, this method only has something to say about a prior
meta-analysis that was not yet significant:

```{r simulation-na}
prev_sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
new_small <- list(yi = rep(log(0.4), 3), vi = rep(0.01, 3), k = 3)
simulation(prev_sig, new_small, seed = 1)
```

**What is wrong with it.** In the same published comparison that produced
the fail-safe-N numbers above, the simulation method flagged **none of the
80 reviews** as out of date — not one. That is either the sign of an
extremely conservative method that will almost never generate a false
alarm, or the sign of a method that is misspecified for the way it was
applied, and the published literature does not settle which. It is exactly
the kind of question this package's backtesting engine exists to make
answerable with data instead of leaving it as a footnote.

## Sources

- Ioannidis JPA, Lau J. Evolution of treatment effects over time: empirical
  insight from recursive cumulative metaanalyses. *Proceedings of the
  National Academy of Sciences*, 2001.
- Shojania KG et al. How quickly do systematic reviews go out of date? A
  survival analysis. *Annals of Internal Medicine*, 2007.
- Barrowman NJ et al. Identifying null meta-analyses that are ripe for
  updating. *BMC Medical Research Methodology*, 2003.
- Mullen B, Muellerleile P, Bryant B. Cumulative meta-analysis: a
  consideration of indicators of sufficiency and stability. *Personality and
  Social Psychology Bulletin*, 2001.
- Pattanittum P et al. A comparison of statistical methods for identifying
  out-of-date systematic reviews. *PLoS ONE*, 2012;7(11):e48894 (PMC3502410).
- Becker BJ. Failsafe N or file-drawer number. In *Publication Bias in
  Meta-Analysis*, 2005.
