---
title: "The five detectors and what is wrong with them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The five detectors and what is wrong with them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(staleness)
library(metafor)
```

Five statistical methods for deciding whether a meta-analysis needs updating
were published between 2003 and 2014. None of them ever received a reusable
software implementation, so nobody has ever been able to run all five against
the same body of evidence and see where they agree. This package implements
all five as pure functions with a common signature — `(prev, new_ma, ...)` in,
a `staleness_verdict` out — precisely so that can finally be done (see
`vignette("backtesting", package = "staleness")` for what happens when it is).

This vignette is deliberately unflattering. Each section below states a
detector's formula, where it comes from, and the specific way the published
literature — or this package's own backtesting — shows it can mislead you.
Implementing a method faithfully is not the same as endorsing it.

## `rcma()`: recursive cumulative meta-analysis

```
ratio = theta_updated / theta_prior      signal if ratio <= 0.5 or ratio >= 1.5
```

The idea, due to Ioannidis and Lau's work on how treatment effects evolve as
evidence accumulates (*Proceedings of the National Academy of Sciences*,
2001), is the simplest of the five: refit the meta-analysis after each new
study and watch how far the pooled estimate moves. A move of 50% or more in
either direction is the signal.

```{r rcma-signal}
prev   <- rma(yi = rep(log(0.50), 4), vi = rep(0.04, 4), measure = "RR")
new_ma <- rma(yi = rep(log(0.20), 4), vi = rep(0.04, 4), measure = "RR")
rcma(prev, new_ma)
```

**What is wrong with it.** The ratio is only well-behaved for measures that
are themselves ratios (risk ratio, odds ratio, hazard ratio): those live on a
log scale where `exp(a - b)` is a well-defined, symmetric ratio regardless of
how small the effect is. For difference measures — a mean difference, a raw
risk difference — the ratio is undefined once the prior effect is close to
zero: a prior estimate of 0.02 moving to 0.40 produces a ratio of 20, an
artefact of the denominator, not a real twenty-fold change. The method as
published does not address this case at all.

```{r rcma-na}
prev_md <- rma(yi = rep(0.02, 4), vi = rep(0.25, 4), measure = "MD")
new_md  <- rma(yi = rep(0.40, 4), vi = rep(0.25, 4), measure = "MD")
rcma(prev_md, new_md)
```

This package's answer is to read the measure straight off the `rma` object
and, for difference measures with a prior effect indistinguishable from zero,
return `not_applicable` with the reason attached rather than a number that
looks plausible and is not.

## `ottawa()`: the Ottawa method

Two quantitative signals: a change in statistical significance (using
**p < 0.04**, not the conventional 0.05 — this is the threshold in the
original method, not a typo), and a change in the pooled effect of at least
50%, computed the same way `rcma()` computes it. The method is due to Shojania
and colleagues' survival analysis of how quickly systematic reviews go out of
date (*Annals of Internal Medicine*, 2007), the same paper that established
the headline numbers this package's introduction vignette opens with.

```{r ottawa-signal}
prev_o <- rma(yi = c(0.02, -0.01, 0.03, -0.02), vi = rep(0.05, 4), measure = "RR")
new_o  <- rma(yi = rep(log(0.40), 8), vi = rep(0.01, 8), measure = "RR")
ottawa(prev_o, new_o)
```

The original method also lists four *qualitative* signals: opposing
findings, evidence of substantial harm, a new superior therapy, and a change
in the balance of effectiveness. **This package does not infer any of
them.** They are accepted as a character vector supplied by the analyst
(`qualitative = "substantial harm reported"`, for instance) and reported
separately from the two quantitative signals. An algorithm that claimed to
judge "substantial harm" from effect sizes alone would be a much bigger
promise than the data can support, and would spend the credibility of the
whole package to make it.

**Declared: `ottawa` contains `rcma` verbatim.** The effect-change half of
the Ottawa method *is* `rcma()`'s entire rule — the same ratio, computed by
the same internal function, against the same 0.5 and 1.5 thresholds. At
default parameters, therefore:

```
rcma(prev, new_ma) == "out_of_date"   implies   ottawa(prev, new_ma) == "out_of_date"
```

always, by construction. Verified over 400 random prior/updated pairs:
`rcma` fired 63 times, `ottawa` fired 116 times, and every one of `rcma`'s 63
was among `ottawa`'s — zero disjoint firings, as the algebra requires.

This is faithful to the published methods and is not a defect in either: the
Ottawa method genuinely adopts the rCMA effect-change criterion and adds a
significance criterion and four qualitative ones on top of it. It is declared
here for the same reason `CONTAMINATED_PAIRS` exists. This package goes to
some trouble to make sure nobody mistakes a structural identity between a
detector and a truth definition for an empirical finding; a structural
dependency between two *detectors* deserves the same treatment. The headline
question this package exists to re-ask — the five methods agreed at
Kappa = 0.14 on 80 reviews — involves ten detector pairs, and one of them is
deterministic. Anyone recomputing that agreement should be counting nine.

## `barrowman()`: the participant-ratio method

```
q = m / n            n = (Z_crit^2 * N) / Z^2            signal if q > 1
```

`m` is the number of participants contributed by the new studies; `N` is the
total in the prior meta-analysis; `Z` is that meta-analysis's own test
statistic; `Z_crit` is 1.96. `n` is a back-calculation: how many new
participants, added at the same average effect and variance as the existing
evidence, would it take to push the pooled result across the significance
threshold. `q > 1` means the new evidence already supplied more than that.

```{r barrowman-signal}
prev_b <- rma(yi = c(0.10, -0.05, 0.08, -0.02), vi = rep(0.05, 4), measure = "MD")
z <- prev_b$zval
n_req <- (1.96^2 * 400) / z^2
barrowman(prev_b, n_prev = 400, n_new = ceiling(n_req * 1.5))
barrowman(prev_b, n_prev = 400, n_new = floor(n_req * 0.5))
```

**What is wrong with it.** The method, proposed by Barrowman and colleagues
as a way to identify null meta-analyses ripe for updating (*BMC Medical
Research Methodology*, 2003), only makes sense when the prior meta-analysis
was *not* statistically significant — `n` is derived from how far the prior
result sits from the significance boundary, and that derivation has no
answer when the prior result was already on the far side of it. This
restricts the method to a narrow slice of published meta-analyses. When it
does not apply, `barrowman()` returns `not_applicable`, never `current`:
treating "the method has nothing to say" as "the review is fine" would hand
this detector an unearned advantage in any calibration exercise, since it
would then never be wrong on exactly the cases it declined to judge.

```{r barrowman-na}
sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
barrowman(sig, n_prev = 400, n_new = 900)
```

## `sufficiency()`: sufficiency and stability

```
index = N_fs / (5k + 10)      stability = has the cumulative pooled effect
                                          stopped moving?
```

`N_fs` is Rosenthal's fail-safe N: an estimate of how many unpublished null
studies it would take to erase a significant pooled effect. A review is
*sufficient* when the index exceeds 1 — Rosenthal's own rule of thumb for a
result being robust to unpublished null findings. It is *stable* when the
cumulative pooled effect, accumulated study by study in publication order,
has no more movement left in it than the same studies would produce in a
random order. The published rule states that as the slope of a regression
across the cumulative effects; this package tests a different statistic,
which is a **declared departure** — the largest of the package's adaptations
— and the next-but-one subsection sets out exactly what was changed, why the
published statistic was measured and rejected, and what the substitute does
instead.

The method is due to Mullen, Muellerleile and Bryant's treatment of
cumulative meta-analysis indicators (*Personality and Social Psychology
Bulletin*, 2001). This package follows the specific two-snapshot form of it
used in the comparative evaluation of Pattanittum and colleagues (*PLoS
ONE*, 2012), because that is the only source that defines sufficiency and
stability as a comparison between a *prior* meta-analysis and an *updated*
one, rather than as a property of a single, static body of studies.

**The rule, and why it is easy to get backwards.** A review counts as
out of date only when it is **both sufficient and unstable at once**:

```{r sufficiency-out}
yi <- seq(log(0.9), log(0.2), length.out = 20)
vi <- rep(0.01, 20)
drifting <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(drifting, drifting)
```

Insufficiency by itself — an index at or below 1 — is **never** grounds for
"out of date", no matter how unstable the cumulative effect looks:

```{r sufficiency-insufficient}
yi <- rep(log(0.99), 6)
vi <- rep(0.04, 6)
weak <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(weak, weak)
```

This is the primary source's own reading, confirmed against its data, not
just its prose: of the 80 reviews in the Pattanittum et al. comparison, all
80 had a fail-safe ratio below 1 (insufficient), and for exactly that reason
the method could not flag a single one of them as out of date. Had the
opposite direction been correct — insufficiency implying "out of date" — that
same fact about the 80 reviews would have produced the opposite result: all
80 flagged.

The two halves also read from different bodies of evidence, and that is
deliberate, not an oversight: sufficiency — the fail-safe N and the `k` in
`5k + 10` — is computed from the meta-analysis **as it stood when
previously published**; stability is computed from the cumulative series of
the **updated** meta-analysis, because that is where drift from new studies
would actually show up. See section 4.4 of the package's design document for
the full paper trail behind this, including the case that would silently
give the wrong verdict if both halves were read from the same snapshot.

### Declared adaptation: how stability is actually tested

Every other adaptation in this package is documented at length — `rcma()`'s
`not_applicable` rule for difference measures, `ottawa()`'s refusal to infer
qualitative signals, the prev/updated split above, the inverted direction of
the out-of-date criterion. This one is the last of them, and it matters most,
because the backtesting results will be read as findings about the published
method.

**What the source says.** Pattanittum et al. (2012, Table 1), reproducing
Mullen et al.'s rule, states the out-of-date criterion as *"failsafe ratio >1
and absolute slope of the linear regression >0"*, where the regression is
*"fitted across the cumulative treatment effects ... versus information
increment"*.

**Why it cannot be implemented as written.** On continuous data the
least-squares slope of a cumulative series is never exactly zero. Taken
literally, `|slope| > 0` is true of essentially every real meta-analysis, so
the stability indicator would carry no information at all and the method
would reduce to "sufficient ⇒ out of date". Some rule for deciding when a
slope is *large enough* has to stand in for it.

Three substitutes were implemented and measured. The first two failed, and
the numbers are given here rather than summarised, because the point of this
package is to adjudicate these methods with data.

**Attempt 1 — the OLS t-test. Invalid.** The obvious substitute is the
t-test on the slope that `lm()` reports, treating `|slope| > 0` as "slope
significantly different from zero". That test is invalid here, and
measurably so. A cumulative weighted mean is near-perfectly autocorrelated by
construction, and it converges on the pooled effect by the law of large
numbers whether or not anything is changing, so the OLS null distribution
does not apply: the test detects *convergence* and reports *instability*.
Measured over 300 samples of evidence that genuinely does not change — 20
prior studies and 10 new ones drawn from one distribution — it returned
`out_of_date` **209 times**. `rcma()` and `ottawa()` returned it zero times
on the same 300 samples. At the degenerate limit it was worse: twelve
byte-identical studies give a cumulative series spanning 2.2e-16, and the fit
obligingly returned a slope of -7.4e-17 with p = 0.014, i.e. "out of date".

**Attempt 2 — permute study order, keep the slope. Calibrated on average;
blind exactly where it mattered.** Replacing the OLS null with a
**permutation test over study order** — compare the observed slope with the
slopes of the same studies reshuffled 999 times — fixed the headline number:
16 firings in 300 on the same no-change experiment, the nominal 5%. It also
rests on a defensible principle, that study order is exchangeable when
nothing is changing. Two independent measurements then showed it does not
work:

* **No power against late change.** With 20 prior studies at RR 0.5 and 10
  new ones at RR 0.30, it fired on **1 of 200** samples, and on **0 of 200**
  once the new studies moved to RR 0.15 or beyond — power that *fell* as the
  change grew. On a 20-study series sitting at RR 0.5 and then dropping to
  RR 0.05, it fired when the drop came after study 2, 5, 8 or 10 and was
  silent when it came after study 12, 15 or 18. The mechanism is that the
  slope of a cumulative series is dominated by its first few points, where
  almost no information has accumulated and the series swings hardest: a
  permuted order that happens to open on an outlying study produces a larger
  slope than a genuine cliff at study 15 ever does. A mature review updated
  with a smaller batch of new studies is this package's canonical case, and
  it is precisely the case the detector could not see.
* **False alarms under time-correlated heteroscedasticity.** With no drift at
  all but variance falling over time — early small trials, later large ones,
  which is what a real evidence stream looks like — it fired on **83 of 300**
  samples (28%), and on a schedule of twenty small trials followed by ten
  large ones, on **127 of 300** (42%). Study order is exchangeable under no
  change only if the *variances* are exchangeable too, and in a real series
  they are not: the cumulative estimate wanders while only small trials have
  accumulated and then settles when the large ones arrive, which looks like
  drift, while permuted orders put the large trials early half the time and
  never show it.

**A fidelity fix that is not a repair.** The regression above was fitted
against the study index `1, 2, 3, ...`, not against accumulated information
`cumsum(1 / vi)`, which is what the source's *"versus information increment"*
specifies and what trial-sequential-analysis methods use. That was a genuine
infidelity and it is now fixed — `detail$slope` is fitted on the information
axis. It repairs **nothing**, and that was measured rather than assumed: when
study variances are equal, `cumsum(1 / vi)` is an affine function of the
index, so every slope in the permutation — observed and null alike — is
rescaled by the same constant and the p-value is identical *to the last bit*.
All three failing measurements above were unchanged by it, and the 28%
false-alarm rate moved only to 25%. A plausible mechanism and fidelity to the
source were not enough; the measurement is what settled it.

**Attempt 3 — change the statistic. What runs.** The slope over a cumulative
series is simply a poor instrument for the question the method is asking. It
is replaced with the **largest standardised movement the cumulative series
has left in it**: for each split point *m*, the fixed-effect pooled estimate
of the first *m* studies is compared with the pooled estimate of the rest,
divided by the standard error of that difference, and the statistic is the
maximum absolute value over *m*.

```
I_m = sum of 1/v_i over the first m studies      (information accumulated)

        theta_hat(1..m)  -  theta_hat(m+1..k)
Z_m = ------------------------------------------      T = max |Z_m|
        sqrt( 1/I_m  +  1/(I_k - I_m) )                    m
```

This still asks the method's own question, and it is monotone in the same
quantity the published slope is a proxy for: the movement remaining in the
cumulative curve after step *m* is exactly `1 - I_m/I_k` times the difference
`Z_m` standardises. What it adds is that it is **pivotal** — under no drift
each `Z_m` is standard normal whatever the study variances are — so the
statistic carries no imprint of the variance schedule and the permutation
null becomes valid again. Everything around it is unchanged: `n_perm` (999)
order permutations, the two-sided `(1 + count) / (n_perm + 1)` estimator so
the p-value is never exactly zero, reported as `p_stability` and compared
against `alpha_stability` (default 0.05); a fixed seed; the caller's random
stream saved and restored; and a cumulative series that is constant to within
floating-point rounding short-circuited as maximally stable before anything
is computed.

Measured against the same four experiments, with the two predecessors for
comparison:

| Experiment (no drift = false alarms, drift = power) | OLS *t* | Perm + slope | Perm + `max Z_m` |
|---|---|---|---|
| No drift, 20 + 10 studies, equal variances (300) | 209 | 16 | **15** (5.0%) |
| Power, 10 new studies at RR 0.40 / 0.30 / 0.15 / 0.02 (200 each) | — | 10 / 1 / 0 / 0 | **200 / 200 / 200 / 200** |
| Shift-position scan, drop after study 2 / 5 / 8 / 10 / 12 / 15 / 18 | — | fires / fires / fires / fires / — / — / — | **fires at all seven** |
| No drift, variance falling over time (300) | — | 83 (28%) | **16** (5.3%) |
| No drift, variance rising over time (300) | — | 0 (no power left) | **20** (6.7%) |
| No drift, 20 small then 10 large trials (300) | — | 127 (42%) | **19** (6.3%) |

Calibration stays between 4.0% and 6.7% across nine variance and
heterogeneity regimes at `k = 30` — including between-study heterogeneity up
to `tau = 0.20` and lognormally distributed variances — and reads 5.0% at
`k = 8` and 4.5% at `k = 60`. The one exception is
`k = 5`, the minimum this detector accepts, where only four split points and
120 distinct orderings exist and the test is **conservative** (1.0% rather
than 5%) — it under-fires there rather than over-fires, which is the safe
direction for a false-alarm-sensitive instrument but does mean the detector
has little power at the bottom of its range.

Two other repairs were implemented and measured before this one was adopted,
and are recorded because negative results are results. **Permuting
standardised residuals** while holding the variance schedule fixed does fix
the heteroscedasticity false alarms (83/300 → 15/300) but cannot touch the
power problem at all: when variances are equal it is *identical* to permuting
study order, so the shift-position scan was unchanged. **Weighting the
regression** by accumulated information restores a great deal of power
(1/200 → 200/200 at RR 0.30, and the scan reaches study 12) but leaves the
heteroscedastic false-alarm rate at 42/300 (14%) and 78/300 (26%) on the two
schedules above — better, still broken. Only replacing the statistic cleared
both bars.

The `slope` itself is still reported in `detail`, now on the information axis
the source specifies, so anyone who wants to apply the literal published rule
can read it straight off the verdict. It no longer decides anything.

**What is still wrong with the statistic that runs.** Three things, all
measured, none of them hidden — a detector with a disclosed blind spot is
honest, one with an undisclosed one is not.

* **Conservative at `k = 5`**, as above: 1.0% rather than 5%, so a `current`
  verdict from a five-study update is weak evidence of stability.
* **A single strongly discordant study, arriving first or last, can carry
  the statistic.** The maximum runs over every split point, including the two
  that isolate one study. A study 20 standard errors from the rest, at either
  end of the series, reaches a permutation p-value of about `2 / k` on its
  own: not significant at `k = 20` (p = 0.12), significant from around
  `k = 35` (p = 0.045 at `k = 40`). That is arguably correct — such a study
  does move the pooled estimate, and the ordering really is extreme among all
  orderings — but it means an `out_of_date` verdict on a long series can rest
  on one outlier rather than a trend. `detail$split` distinguishes the two: a
  split at `k - 1` is one study, a split in the middle is a body of them.
* **Power falls away under strong between-study heterogeneity.** Against 20
  prior studies at RR 0.5 and 10 new at RR 0.30: 200/200 at `tau = 0` and
  `tau = 0.10`, 199/200 at `tau = 0.20`, 160/200 at `tau = 0.35`. The
  permutation null is built from the same heterogeneous studies, so
  calibration holds throughout; it is power, not the false-alarm rate, that
  degrades. The detector is at its sharpest where the size of the change is
  large relative to the spread between studies — against 20 prior studies at
  RR 0.5, ten new ones at RR 0.50 fire on 8/200 samples (the nominal rate),
  RR 0.47 on 117/200, RR 0.45 on 197/200, and RR 0.42 and beyond on 200/200.

**What is wrong with it, independent of any of the above.** Rosenthal's
fail-safe N has been discredited as a measure of publication bias since
Becker (2005) — it answers a narrower and less useful question than it is
usually taken to answer, and better alternatives (trim-and-fill, funnel-plot
regression) have existed for years. `sufficiency()` implements it
anyway, faithfully, exactly as the published method specifies it, because
this package's purpose is not to pre-judge which of the five methods
deserves to survive contact with data. If Rosenthal's N is as weak a
foundation as the critique suggests, the backtesting engine will show it —
that is a fact to be established empirically, in
`vignette("backtesting", package = "staleness")`, and not decided in advance
by leaving the method out.

## `simulation()`: prospective power

Ten thousand replicate "new studies" are simulated using the parameters of
the recently published literature, added to the prior meta-analysis, and the
fraction of replicates in which the pooled result reaches significance is
recorded as a power estimate. A power at or above 80% is the signal.

```{r simulation-signal}
prev_sim <- rma(yi = c(0.30, 0.28, 0.32, 0.29), vi = rep(0.16, 4), measure = "MD")
new_ev   <- list(yi = c(0.30, 0.31, 0.29), vi = rep(0.005, 3), k = 3)
simulation(prev_sim, new_ev, seed = 3)
```

Like `barrowman()`, this method only has something to say about a prior
meta-analysis that was not yet significant:

```{r simulation-na}
prev_sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
new_small <- list(yi = rep(log(0.4), 3), vi = rep(0.01, 3), k = 3)
simulation(prev_sig, new_small, seed = 1)
```

**What is wrong with it.** In the same published comparison that produced
the fail-safe-N numbers above, the simulation method flagged **none of the
80 reviews** as out of date — not one. That is either the sign of an
extremely conservative method that will almost never generate a false
alarm, or the sign of a method that is misspecified for the way it was
applied, and the published literature does not settle which. It is exactly
the kind of question this package's backtesting engine exists to make
answerable with data instead of leaving it as a footnote.

## Sources

- Ioannidis JPA, Lau J. Evolution of treatment effects over time: empirical
  insight from recursive cumulative metaanalyses. *Proceedings of the
  National Academy of Sciences*, 2001.
- Shojania KG et al. How quickly do systematic reviews go out of date? A
  survival analysis. *Annals of Internal Medicine*, 2007.
- Barrowman NJ et al. Identifying null meta-analyses that are ripe for
  updating. *BMC Medical Research Methodology*, 2003.
- Mullen B, Muellerleile P, Bryant B. Cumulative meta-analysis: a
  consideration of indicators of sufficiency and stability. *Personality and
  Social Psychology Bulletin*, 2001.
- Pattanittum P et al. A comparison of statistical methods for identifying
  out-of-date systematic reviews. *PLoS ONE*, 2012;7(11):e48894 (PMC3502410).
- Becker BJ. Failsafe N or file-drawer number. In *Publication Bias in
  Meta-Analysis*, 2005.
