---
title: "The five detectors and what is wrong with them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The five detectors and what is wrong with them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(staleness)
library(metafor)
```

Five statistical methods for deciding whether a meta-analysis needs updating
were published between 1999 and 2007. None of them ever received a reusable
software implementation — a search of all of CRAN finds the building blocks
(`metafor` for cumulative meta-analyses and the fail-safe N) but none of the
five detectors — so nobody has ever been able to run all five against the same
body of evidence and see where they agree. This package implements
all five as pure functions with a common signature — `(prev, new_ma, ...)` in,
a `staleness_verdict` out — precisely so that can finally be done (see
`vignette("backtesting", package = "staleness")` for what happens when it is).

This vignette is deliberately unflattering. Each section below states a
detector's formula, where it comes from, and the specific way the published
literature — or this package's own backtesting — shows it can mislead you.
Implementing a method faithfully is not the same as endorsing it.

## `rcma()`: recursive cumulative meta-analysis

```
ratio = theta_updated / theta_prior      signal if ratio <= 0.5 or ratio >= 1.5
```

The idea, due to Ioannidis and Lau's work on how treatment effects evolve as
evidence accumulates (*Proceedings of the National Academy of Sciences*,
2001), is the simplest of the five: refit the meta-analysis after each new
study and watch how far the pooled estimate moves. A move of 50% or more in
either direction is the signal.

```{r rcma-signal}
prev   <- rma(yi = rep(log(0.50), 4), vi = rep(0.04, 4), measure = "RR")
new_ma <- rma(yi = rep(log(0.20), 4), vi = rep(0.04, 4), measure = "RR")
rcma(prev, new_ma)
```

**What is wrong with it.** The ratio is only well-behaved for measures that
are themselves ratios (risk ratio, odds ratio, hazard ratio): those live on a
log scale where `exp(a - b)` is a well-defined, symmetric ratio regardless of
how small the effect is. For difference measures — a mean difference, a raw
risk difference — the ratio is undefined once the prior effect is close to
zero: a prior estimate of 0.02 moving to 0.40 produces a ratio of 20, an
artefact of the denominator, not a real twenty-fold change. The method as
published does not address this case at all.

```{r rcma-na}
prev_md <- rma(yi = rep(0.02, 4), vi = rep(0.25, 4), measure = "MD")
new_md  <- rma(yi = rep(0.40, 4), vi = rep(0.25, 4), measure = "MD")
rcma(prev_md, new_md)
```

This package's answer is to read the measure straight off the `rma` object
and, for difference measures with a prior effect indistinguishable from zero,
return `not_applicable` with the reason attached rather than a number that
looks plausible and is not.

## `ottawa()`: the Ottawa method

Two quantitative signals: a change in statistical significance (using
**p < 0.04**, not the conventional 0.05 — this is the threshold in the
original method, not a typo), and a change in the pooled effect of at least
50%, computed the same way `rcma()` computes it. The method is due to Shojania
and colleagues' survival analysis of how quickly systematic reviews go out of
date (*Annals of Internal Medicine*, 2007), the same paper that established
the headline numbers this package's introduction vignette opens with.

```{r ottawa-signal}
prev_o <- rma(yi = c(0.02, -0.01, 0.03, -0.02), vi = rep(0.05, 4), measure = "RR")
new_o  <- rma(yi = rep(log(0.40), 8), vi = rep(0.01, 8), measure = "RR")
ottawa(prev_o, new_o)
```

The original method also lists four *qualitative* signals: opposing
findings, evidence of substantial harm, a new superior therapy, and a change
in the balance of effectiveness. **This package does not infer any of
them.** They are accepted as a character vector supplied by the analyst
(`qualitative = "substantial harm reported"`, for instance) and reported
separately from the two quantitative signals. An algorithm that claimed to
judge "substantial harm" from effect sizes alone would be a much bigger
promise than the data can support, and would spend the credibility of the
whole package to make it.

**Corrected: `ottawa` does *not* contain `rcma`.** Earlier versions of this
vignette said it did, on the reasonable-looking grounds that both compare an
updated effect to a prior one at 0.5 and 1.5. They compare different
quantities. `rcma()` uses the ratio of the effects; the Ottawa method uses the
ratio of the relative risk *reductions*:

```
RRR ratio = (1 - RR_updated) / (1 - RR_previous)      signal if <= 0.5 or >= 1.5
```

That is the definition its published application states (Pattanittum et al.,
2012, Table 1), and this package now follows it. For mean differences the
source defers to the rCMA rule, so there the two do coincide.

The difference is not academic. That study's Appendix S3 publishes the ten
reviews with the largest Ottawa indicator, with both pooled effects and the
ratio the authors computed. Reproduced in this package's test suite: **all ten
fire on the RRR ratio, and none of them fires on the ratio of the risk
ratios.** A review moving from RR 0.995 to RR 0.848 barely moves the risk
ratio — 0.85, comfortably inside the band — while its risk reduction goes from
0.5% to 15%.

**What is wrong with it.** The denominator `1 - RR_previous` goes to zero as
the prior effect approaches no effect, so the ratio explodes exactly where the
method is meant to be used: on meta-analyses whose result is not yet
significant. Measured on evidence containing no change whatsoever, the effect
signal fires on **64% of samples** under a null effect, and on **0%** once the
effect is real and precise.

That is not a quirk of this implementation; it follows from the criterion, and
it explains the published comparison. Ottawa flagged 34 of the 80 reviews where
recursive CMA and Barrowman flagged 7 each — and every one of those 80 was a
null meta-analysis by inclusion criterion. The headline figure this package
exists to re-ask, Kappa = 0.14, was computed over the three methods that
discriminated at all, one of which was firing most of the time for structural
reasons.

## `barrowman()`: the participant-ratio method

```
q = m / n            n = (Z_crit^2 * N) / Z^2            signal if q > 1
```

`m` is the number of participants contributed by the new studies; `N` is the
total in the prior meta-analysis; `Z` is that meta-analysis's own test
statistic; `Z_crit` is 1.96. `n` is a back-calculation: how many new
participants, added at the same average effect and variance as the existing
evidence, would it take to push the pooled result across the significance
threshold. `q > 1` means the new evidence already supplied more than that.

```{r barrowman-signal}
prev_b <- rma(yi = c(0.10, -0.05, 0.08, -0.02), vi = rep(0.05, 4), measure = "MD")
z <- prev_b$zval
n_req <- (1.96^2 * 400) / z^2
barrowman(prev_b, n_prev = 400, n_new = ceiling(n_req * 1.5))
barrowman(prev_b, n_prev = 400, n_new = floor(n_req * 0.5))
```

**What is wrong with it.** The method, proposed by Barrowman and colleagues
as a way to identify null meta-analyses ripe for updating (*BMC Medical
Research Methodology*, 2003), only makes sense when the prior meta-analysis
was *not* statistically significant — `n` is derived from how far the prior
result sits from the significance boundary, and that derivation has no
answer when the prior result was already on the far side of it. This
restricts the method to a narrow slice of published meta-analyses. When it
does not apply, `barrowman()` returns `not_applicable`, never `current`:
treating "the method has nothing to say" as "the review is fine" would hand
this detector an unearned advantage in any calibration exercise, since it
would then never be wrong on exactly the cases it declined to judge.

```{r barrowman-na}
sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
barrowman(sig, n_prev = 400, n_new = 900)
```

## `sufficiency()`: sufficiency and stability

```
index = N_fs / (5k + 10)      stability = has the cumulative pooled effect
                                          stopped moving?
```

`N_fs` is Rosenthal's fail-safe N: an estimate of how many unpublished null
studies it would take to erase a significant pooled effect. A review is
*sufficient* when the index exceeds 1 — Rosenthal's own rule of thumb for a
result being robust to unpublished null findings. It is *stable* when the
cumulative pooled effect, accumulated study by study in publication order,
has no more movement left in it than the same studies would produce in a
random order. The published rule states that as the slope of a regression
across the cumulative effects; this package tests a different statistic,
which is a **declared departure** — the largest of the package's adaptations
— and the next-but-one subsection sets out exactly what was changed, why the
published statistic was measured and rejected, and what the substitute does
instead.

The method is due to Mullen, Muellerleile and Bryant's treatment of
cumulative meta-analysis indicators (*Personality and Social Psychology
Bulletin*, 2001). This package follows the specific two-snapshot form of it
used in the comparative evaluation of Pattanittum and colleagues (*PLoS
ONE*, 2012), because that is the only source that defines sufficiency and
stability as a comparison between a *prior* meta-analysis and an *updated*
one, rather than as a property of a single, static body of studies.

**The rule, and why it is easy to get backwards.** A review counts as
out of date only when it is **both sufficient and unstable at once**:

```{r sufficiency-out}
yi <- seq(log(0.9), log(0.2), length.out = 20)
vi <- rep(0.01, 20)
drifting <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(drifting, drifting)
```

Insufficiency by itself — an index at or below 1 — is **never** grounds for
"out of date", no matter how unstable the cumulative effect looks:

```{r sufficiency-insufficient}
yi <- rep(log(0.99), 6)
vi <- rep(0.04, 6)
weak <- rma(yi = yi, vi = vi, measure = "RR")
sufficiency(weak, weak)
```

This is the primary source's own reading, confirmed against its data, not
just its prose: of the 80 reviews in the Pattanittum et al. comparison, all
80 had a fail-safe ratio below 1 (insufficient), and for exactly that reason
the method could not flag a single one of them as out of date. Had the
opposite direction been correct — insufficiency implying "out of date" — that
same fact about the 80 reviews would have produced the opposite result: all
80 flagged.

The two halves also read from different bodies of evidence, and that is
deliberate, not an oversight: sufficiency — the fail-safe N and the `k` in
`5k + 10` — is computed from the meta-analysis **as it stood when
previously published**; stability is computed from the cumulative series of
the **updated** meta-analysis, because that is where drift from new studies
would actually show up. See section 4.4 of the package's design document for
the full paper trail behind this, including the case that would silently
give the wrong verdict if both halves were read from the same snapshot.

### Declared adaptation: how stability is actually tested

Every other adaptation in this package is documented at length — `rcma()`'s
`not_applicable` rule for difference measures, `ottawa()`'s refusal to infer
qualitative signals, the prev/updated split above, the inverted direction of
the out-of-date criterion. This one is the last of them, and it matters most,
because the backtesting results will be read as findings about the published
method.

**What the source says.** Pattanittum et al. (2012, Table 1), reproducing
Mullen et al.'s rule, states the out-of-date criterion as *"failsafe ratio >1
and absolute slope of the linear regression >0"*, where the regression is
*"fitted across the cumulative treatment effects ... versus information
increment"*.

**Why it cannot be implemented as written.** On continuous data the
least-squares slope of a cumulative series is never exactly zero. Taken
literally, `|slope| > 0` is true of essentially every real meta-analysis, so
the stability indicator would carry no information at all and the method
would reduce to "sufficient ⇒ out of date". Some rule for deciding when a
slope is *large enough* has to stand in for it.

Three substitutes were implemented and measured. The first two failed, and
the numbers are given here rather than summarised, because the point of this
package is to adjudicate these methods with data.

**Attempt 1 — the OLS t-test. Invalid.** The obvious substitute is the
t-test on the slope that `lm()` reports, treating `|slope| > 0` as "slope
significantly different from zero". That test is invalid here, and
measurably so. A cumulative weighted mean is near-perfectly autocorrelated by
construction, and it converges on the pooled effect by the law of large
numbers whether or not anything is changing, so the OLS null distribution
does not apply: the test detects *convergence* and reports *instability*.
Measured over 300 samples of evidence that genuinely does not change — 20
prior studies and 10 new ones drawn from one distribution — it returned
`out_of_date` **209 times**. `rcma()` and `ottawa()` returned it zero times
on the same 300 samples. At the degenerate limit it was worse: twelve
byte-identical studies give a cumulative series spanning 2.2e-16, and the fit
obligingly returned a slope of -7.4e-17 with p = 0.014, i.e. "out of date".

**Attempt 2 — permute study order, keep the slope. Calibrated on average;
blind exactly where it mattered.** Replacing the OLS null with a
**permutation test over study order** — compare the observed slope with the
slopes of the same studies reshuffled 999 times — fixed the headline number:
16 firings in 300 on the same no-change experiment, the nominal 5%. It also
rests on a defensible principle, that study order is exchangeable when
nothing is changing. Two independent measurements then showed it does not
work:

* **No power against late change.** With 20 prior studies at RR 0.5 and 10
  new ones at RR 0.30, it fired on **1 of 200** samples, and on **0 of 200**
  once the new studies moved to RR 0.15 or beyond — power that *fell* as the
  change grew. On a 20-study series sitting at RR 0.5 and then dropping to
  RR 0.05, it fired when the drop came after study 2, 5, 8 or 10 and was
  silent when it came after study 12, 15 or 18. The mechanism is that the
  slope of a cumulative series is dominated by its first few points, where
  almost no information has accumulated and the series swings hardest: a
  permuted order that happens to open on an outlying study produces a larger
  slope than a genuine cliff at study 15 ever does. A mature review updated
  with a smaller batch of new studies is this package's canonical case, and
  it is precisely the case the detector could not see.
* **False alarms under time-correlated heteroscedasticity.** With no drift at
  all but variance falling over time — early small trials, later large ones,
  which is what a real evidence stream looks like — it fired on **83 of 300**
  samples (28%), and on a schedule of twenty small trials followed by ten
  large ones, on **127 of 300** (42%). Study order is exchangeable under no
  change only if the *variances* are exchangeable too, and in a real series
  they are not: the cumulative estimate wanders while only small trials have
  accumulated and then settles when the large ones arrive, which looks like
  drift, while permuted orders put the large trials early half the time and
  never show it.

**A fidelity fix that is not a repair.** The regression above was fitted
against the study index `1, 2, 3, ...`, not against accumulated information
`cumsum(1 / vi)`, which is what the source's *"versus information increment"*
specifies and what trial-sequential-analysis methods use. That was a genuine
infidelity and it is now fixed — `detail$slope` is fitted on the information
axis. It repairs **nothing**, and that was measured rather than assumed: when
study variances are equal, `cumsum(1 / vi)` is an affine function of the
index, so every slope in the permutation — observed and null alike — is
rescaled by the same constant and the p-value is identical *to the last bit*.
All three failing measurements above were unchanged by it, and the 28%
false-alarm rate moved only to 25%. A plausible mechanism and fidelity to the
source were not enough; the measurement is what settled it.

**Attempt 3 — change the statistic. What runs.** The slope over a cumulative
series is simply a poor instrument for the question the method is asking. It
is replaced with the **largest standardised movement the cumulative series
has left in it**: for each split point *m*, the fixed-effect pooled estimate
of the first *m* studies is compared with the pooled estimate of the rest,
divided by the standard error of that difference, and the statistic is the
maximum absolute value over *m*.

```
I_m = sum of 1/v_i over the first m studies      (information accumulated)

        theta_hat(1..m)  -  theta_hat(m+1..k)
Z_m = ------------------------------------------      T = max |Z_m|
        sqrt( 1/I_m  +  1/(I_k - I_m) )                    m
```

This still asks the method's own question, and it is monotone in the same
quantity the published slope is a proxy for: the movement remaining in the
cumulative curve after step *m* is exactly `1 - I_m/I_k` times the difference
`Z_m` standardises. What it adds is that its dependence on the variance
schedule is **weak instead of structural**. Two things are exactly true: the
statistic is *scale*-pivotal, so rescaling the effects and their variances
together — `yi` to `theta + a(yi - theta)` and `vi` to `a^2 vi` — leaves it
bit-identical, and each `Z_m` is *marginally* standard normal under any
schedule (measured SDs 0.998–1.002 across schedules spanning 650:1). Scaling
`vi` on its own is a different operation and not one the statistic is blind
to: it divides every `Z_m` by `sqrt(c)`, leaving the *p*-value unchanged
because the permuted statistics rescale with it. What is
**not** true — and an earlier draft of this vignette claimed it — is that the
statistic therefore carries no imprint of the schedule at all. `max_m |Z_m|`
is the discrete supremum of a standardised Brownian bridge sampled at the
information fractions `I_m/I_k`, and those fractions are a function of the
schedule. Null quantiles at `k = 30` over 60,000 draws:

| schedule (variance) | ratio | median | 95th pct |
|---|---|---|---|
| flat | 1:1 | 1.817 | 2.881 |
| V-shaped, big–small–big | 16:1 | 1.923 | 2.950 |
| Λ-shaped, small–big–small | 16:1 | 1.599 | 2.732 |
| geometric decay | 646:1 | 1.930 | 2.956 |

The V and Λ nulls differ with a Kolmogorov–Smirnov `D` of 0.25, and the
flat-schedule 95th percentile delivers 6.1% under V and 3.4% under Λ. So the
case for keeping the order-permutation null is that the residual dependence
is small — quantiles move a few percent where the slope statistic moved by
orders of magnitude — not that it is zero. What that costs is measured below.
Everything around it is unchanged: `n_perm` (999)
order permutations, the two-sided `(1 + count) / (n_perm + 1)` estimator so
the p-value is never exactly zero, reported as `p_stability` and compared
against `alpha_stability` (default 0.05); a fixed seed; the caller's random
stream saved and restored; and a cumulative series that is constant to within
floating-point rounding short-circuited as maximally stable before anything
is computed.

Measured against the same four experiments, with the two predecessors for
comparison:

| Experiment (no drift = false alarms, drift = power) | OLS *t* | Perm + slope | Perm + `max Z_m` |
|---|---|---|---|
| No drift, 20 + 10 studies, equal variances (300) | 209 | 16 | **15** (5.0%) |
| Power, 10 new studies at RR 0.40 / 0.30 / 0.15 / 0.02 (200 each) | — | 10 / 1 / 0 / 0 | **200 / 200 / 200 / 200** |
| Shift-position scan, drop after study 2 / 5 / 8 / 10 / 12 / 15 / 18 | — | fires / fires / fires / fires / — / — / — | **fires at all seven** |
| No drift, variance falling over time (300) | — | 83 (28%) | **16** (5.3%) |
| No drift, variance rising over time (300) | — | 0 (no power left) | **20** (6.7%) |
| No drift, 20 small then 10 large trials, 50:1 (300) | — | 127 (42%) | **20** (6.7%) |

Every number in the table above is reproducible. `inst/calibration/calibration.R`
regenerates all of them from the seeds the original measurements used, and
reconstructs the replaced statistic as well, so the "before" column can be
re-derived rather than taken on trust:

```
Rscript inst/calibration/calibration.R
```

**Calibration, and where it fails.** The first draft of this vignette said
calibration "stays between 4.0% and 6.7% across nine variance and
heterogeneity regimes". That was true of those nine and false as a general
claim: every one of them was flat, mildly linear or a single step. Measured
over 1000 no-drift samples at `k = 30` (2000 for the worst case), against the
actual permutation test:

| variance schedule | ratio | false alarms |
|---|---|---|
| flat | 1:1 | 5.3% |
| linear decay (the E4 schedule) | 16:1 | 5.7% |
| `dat.bcg`, real variances in year order | 60:1 | 5.0% |
| `dat.egger2001`, real variances | 90:1 | 3.0% |
| Λ-shaped, small–big–small | 16:1 | 2.4% (conservative) |
| step, 20 small then 10 large trials | 50:1 | 8.1% |
| V-shaped, big–small–big | 16:1 | 8.4% |
| `dat.bcg`, the same variances sorted decreasing | 60:1 | 9.4% |
| geometric decay, `vi = 0.5 * 0.845^j` | 114:1 | 9.3% |
| **geometric decay, `vi = 0.5 * 0.80^j`** | **650:1** | **11.1%** |
| geometric growth, the same reversed | 650:1 | 10.9% |

What matters is not the direction of the trend — growth and decay are equally
bad — but its **smoothness, monotonicity and range**. Irregular real-world
schedules come out nominal; taking `dat.bcg`'s own variances and merely
sorting them into decreasing order moves it from 5.0% to 9.4%, which isolates
the arrangement from the numbers. The measured worst case is about **11%** on
a smooth 650:1 monotone precision ramp, roughly twice nominal. That is a
disclosed limitation rather than a claim of nominality, and it is pinned by an
invariant test so it cannot silently get worse.

Across series length, at a flat schedule: 0.9% at `k = 5`, 1.7% at `k = 6`,
3.4% at `k = 8`, 3.3% at `k = 10`, 5.5% at `k = 12`, and nominal from there —
still half-nominal at twice the minimum `k` this detector accepts.

Two other repairs were implemented and measured before this one was adopted,
and are recorded because negative results are results. **Permuting
standardised residuals** while holding the variance schedule fixed does fix
the heteroscedasticity false alarms (83/300 → 15/300) but cannot touch the
power problem at all: when variances are equal it is *identical* to permuting
study order, so the shift-position scan was unchanged. **Weighting the
regression** by accumulated information restores a great deal of power
(1/200 → 200/200 at RR 0.30, and the scan reaches study 12) but leaves the
heteroscedastic false-alarm rate at 42/300 (14%) and 78/300 (26%) on the two
schedules above — better, still broken. Only replacing the statistic cleared
both bars.

The `slope` itself is still reported in `detail`, now on the information axis
the source specifies, so anyone who wants to apply the literal published rule
can read it straight off the verdict. It no longer decides anything.

**What is still wrong with the statistic that runs.** Four things, all
measured, none of them hidden — a detector with a disclosed blind spot is
honest, one with an undisclosed one is not. Three of them were understated in
the first draft of this vignette and are restated here with the numbers that
falsified the original wording.

* **Conservative well past `k = 5`**: 0.9% at `k = 5` but still only 3.4% at
  `k = 8` and 3.3% at `k = 10`, reaching nominal around `k = 12`. The first
  draft blamed the 120 distinct orderings available at `k = 5`, which cannot
  explain `k = 8` and its 40,320. The mechanism is the next bullet with its
  sign reversed: the maximum is often driven by the split isolating the single
  most extreme study, and a given study lands at an end in `2 / k` of all
  orderings — 40% at `k = 5`, 25% at `k = 8`, 6.7% at `k = 30`. At small `k`
  a large share of permutations reproduce the observed extreme, so the null is
  diffuse and the test under-fires. A `current` verdict from a short update is
  weak evidence of stability.
* **A single discordant study, arriving first or last, can carry the
  statistic — and the threshold is much lower than first reported.** With
  ordinary sampling noise on the other studies, P(fires) over 400 samples:

  | `k` | no outlier | `D` = 3 SE | `D` = 5 SE | `D` = 10 SE |
  |---|---|---|---|---|
  | 20 | 0.050 | 0.175 | 0.625 | 0.660 |
  | 25 | 0.037 | 0.205 | 0.885 | 0.938 |
  | 30 | 0.045 | 0.210 | 0.927 | 0.990 |
  | 40 | 0.052 | 0.240 | 0.995 | 0.993 |

  So a **five**-SE discordant study at either end forces `unstable` almost
  always from about `k = 25` — not "twenty SE from about `k = 35`", which is
  what the first draft said. That figure came from a noise-free fixture where
  every other study was byte-identical, which makes the mirror-image ordering
  an exact tie, inflates the permutation count, and lands in the one corner of
  the parameter space where the effect looks mild. The same saturation shows
  in the table: at `D` = 20 SE the rate *falls* back to 0.453 at `k = 20`. A
  five-SE discordant study is an ordinary occurrence in a real review, so at
  moderate `k` this detector is substantially a single-outlier detector.
  Whether that is the right answer is arguable — such a study does move the
  pooled estimate — but it must not be read as evidence of a trend.
  `detail$split` separates the two readings and does so reliably: among the
  firings above it pointed at `k - 1`, the single-study split, in 97–99% of
  cases.
* **A smooth, strongly monotone variance schedule runs anti-conservative**, to
  about 11% against a nominal 5%, per the calibration table above.
* **Power falls away under strong between-study heterogeneity.** Against 20
  prior studies at RR 0.5 and 10 new at RR 0.30: 200/200 at `tau = 0` and
  `tau = 0.10`, 199/200 at `tau = 0.20`, 160/200 at `tau = 0.35`. The
  permutation null is built from the same heterogeneous studies, so
  calibration holds throughout; it is power, not the false-alarm rate, that
  degrades. The detector is at its sharpest where the size of the change is
  large relative to the spread between studies — against 20 prior studies at
  RR 0.5, ten new ones at RR 0.50 fire on 8/200 samples (the nominal rate),
  RR 0.47 on 117/200, RR 0.45 on 197/200, and RR 0.42 and beyond on 200/200.

**What is wrong with it, independent of any of the above.** Rosenthal's
fail-safe N has been discredited as a measure of publication bias since
Becker (2005) — it answers a narrower and less useful question than it is
usually taken to answer, and better alternatives (trim-and-fill, funnel-plot
regression) have existed for years. `sufficiency()` implements it
anyway, faithfully, exactly as the published method specifies it, because
this package's purpose is not to pre-judge which of the five methods
deserves to survive contact with data. If Rosenthal's N is as weak a
foundation as the critique suggests, the backtesting engine will show it —
that is a fact to be established empirically, in
`vignette("backtesting", package = "staleness")`, and not decided in advance
by leaving the method out.

## `simulation()`: prospective power

Ten thousand replicate "new studies" are simulated using the parameters of
the recently published literature, added to the prior meta-analysis, and the
fraction of replicates in which the pooled result reaches significance is
recorded as a power estimate. A power at or above 80% is the signal.

```{r simulation-signal}
prev_sim <- rma(yi = c(0.30, 0.28, 0.32, 0.29), vi = rep(0.16, 4), measure = "MD")
new_ev   <- list(yi = c(0.30, 0.31, 0.29), vi = rep(0.005, 3), k = 3)
simulation(prev_sim, new_ev, seed = 3)
```

Like `barrowman()`, this method only has something to say about a prior
meta-analysis that was not yet significant:

```{r simulation-na}
prev_sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
new_small <- list(yi = rep(log(0.4), 3), vi = rep(0.01, 3), k = 3)
simulation(prev_sig, new_small, seed = 1)
```

**What is wrong with it.** In the same published comparison that produced
the fail-safe-N numbers above, the simulation method flagged **none of the
80 reviews** as out of date — not one. That is either the sign of an
extremely conservative method that will almost never generate a false
alarm, or the sign of a method that is misspecified for the way it was
applied, and the published literature does not settle which. It is exactly
the kind of question this package's backtesting engine exists to make
answerable with data instead of leaving it as a footnote.

## Sources

- Ioannidis JPA, Lau J. Evolution of treatment effects over time: empirical
  insight from recursive cumulative metaanalyses. *Proceedings of the
  National Academy of Sciences*, 2001.
- Shojania KG et al. How quickly do systematic reviews go out of date? A
  survival analysis. *Annals of Internal Medicine*, 2007.
- Barrowman NJ et al. Identifying null meta-analyses that are ripe for
  updating. *BMC Medical Research Methodology*, 2003.
- Mullen B, Muellerleile P, Bryant B. Cumulative meta-analysis: a
  consideration of indicators of sufficiency and stability. *Personality and
  Social Psychology Bulletin*, 2001.
- Pattanittum P et al. A comparison of statistical methods for identifying
  out-of-date systematic reviews. *PLoS ONE*, 2012;7(11):e48894 (PMC3502410).
- Becker BJ. Failsafe N or file-drawer number. In *Publication Bias in
  Meta-Analysis*, 2005.
