---
title: "Getting started with staleness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with staleness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A meta-analysis is a photograph, not a standing fact. Roughly a quarter of
systematic reviews are out of date within two years of publication, and half
within five and a half. This package asks, for one meta-analysis and the
evidence published since: is it still current?

```{r setup}
library(staleness)
library(metafor)

dat <- metadat::dat.bcg
es  <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg,
              data = dat)
ma  <- rma(yi, vi, data = es)
stream <- evidence_stream(ma, date = es$year, study_id = seq_along(es$year))
stream
```

`evidence_stream()` is the only function in the package that knows about
dates. Everything downstream of it — every detector, `check_currency()`, the
backtesting engine — works with two frozen snapshots of evidence and never
looks at a clock.

Take the state of knowledge in 1970 and ask what the following decade did to
it:

```{r check}
prev <- snapshot_at(stream, 1970)
new  <- window_between(stream, 1970, 1980)

check_currency(prev, new, methods = c("rcma", "ottawa", "sufficiency"))
```

`snapshot_at()` refits the meta-analysis using only studies published on or
before the cut. `window_between()` collects the studies published in the
following decade. `check_currency()` never sees a date: it receives two
bodies of evidence and answers whether the second one changes the
conclusions of the first.

## Why `not applicable` is not `current`

Three verdicts exist, not two. `barrowman` only applies when the prior
meta-analysis was not statistically significant; asked outside that case it
answers `not applicable`, never `current`.

The distinction is not pedantry. Counting inapplicability as a clean bill of
health would inflate every calibration statistic in this package, and it is
precisely the shortcut that makes a tool look better than it is.

```{r na}
sig <- rma(yi = rep(log(0.4), 8), vi = rep(0.01, 8), measure = "RR")
barrowman(sig, n_prev = 400, n_new = 900)
```

The prior meta-analysis above is already significant (`RR = 0.4` over eight
studies with tight variances), so the participant-ratio method has nothing
to compute: it is not that the review is current, it is that this method has
no opinion here. The same rule holds for every detector: a detector that
cannot answer a question never quietly reports the safe answer.

## Where to go next

`vignette("methods", package = "staleness")` documents each of the five
detectors — its formula, its origin, and the specific way it can mislead
you. `vignette("backtesting", package = "staleness")` covers the harder
question underneath all of this: given that the five methods barely agree
with each other, which one, if any, should you trust? That vignette builds
and interprets a full backtest.
